### 2021-09-14 paper/localpsf.tex
paper cleanup

		Compute $\alpha$, $\mu$, and $\Sigma$ using Algorithm \ref{alg:varhpi_mean_cov}.

@@

Because of the non-negative kernel property, for each $x \in \Omega$, the impulse response $\impulseresponse_x$ is a non-negative function $y \mapsto \impulseresponse_x(y)$. Hence, $\impulseresponse_x$ is a scaled probability distribution. Let $\alpha$ denote the scaling factor, let $\widehat{\impulseresponse}_x := \impulseresponse_x / \alpha(x)$ denote the normalized version of $\impulseresponse_x$, and let $\mu(x)$ and $\Sigma(x)$ denote the mean and covariance of $\widehat{\impulseresponse}_x$, respectively. We make the approximation that the support of $\impulseresponse_x$ is contained within the ellipsoid
\begin{equation}
	\label{eq:support_ellipsoid}
	E_x := \{x' \in \Omega: (x' - \mu(x))^T \Sigma(x)^{-1} (x' - \mu(x)) \le \tau^2\},
\end{equation}
where $\tau$ is a fixed constant. The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations from the mean of the Gaussian distribution with mean $\mu(x)$ and covariance $\Sigma(x)$, i.e., the Gaussian distribution which has the same mean and covariance as $\widehat{\impulseresponse}_x$ (see Figure FIG). The quantity $\tau$ is a parameter that must be chosen by the user of the algorithm. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\impulseresponse_x$. However, the cost of our algorithm will depend on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids $\implies$ lower cost), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. We find that $\tau=3$ yields a reasonable balance between these competing interests, and use $\tau=3$ in our numerical results. The fraction of the ``mass'' of $\impulseresponse_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is conservative and typically far less mass resides in this region.

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
	Let $\alpha:\Omega \rightarrow \mathbb{R}$, $\mu:\Omega \rightarrow \mathbb{R}^d$, and $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}$, be the following functions:
	\begin{equation*}
		\alpha(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
		\mu(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
		\Sigma(x) := \Var(\widehat{\impulseresponse}_x),
	\end{equation*}
	where $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^d\right)$. Let $C$, $\{L^i\}_{i=1}^d$, and ${\{Q^{ij}\}_{i=1}^d}_{j=1}^d$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
		C(x) := 1, \qquad
		L^i(x) := x^i, \qquad
		Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ to denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ to denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
			\alpha =& \left(\Aop^T C\right)^* \\
			\mu^i =& \left(\Aop^T L^i\right)^* / \alpha \\
			\Sigma^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \alpha - \mu^i\cdot \mu^j
		\end{align}
	\end{subequations}
	 for $i=1,\dots, d$, $j=1,\dots,d$.
\end{theorem}

@@

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	{	
		\Input{Operator $\Aop$}
		\Output{$\alpha$, $\mu$, and $\Sigma$}

		\tcp{Compute scaling factor $\alpha$}

@@

		Compute $\alpha = \left(\Aop^T C\right)^*$

		\tcp{Compute mean $\mu$}
		\For{$i=1,2,\dots,d$}{
			Form linear function $L^i(x) = x^i$

			Compute $\mu^i = \left(\Aop^T L^i\right)^* / \alpha$
		}
		\tcp{Compute covariance $\Sigma$}
		\For{$i=1,2,\dots,d$}{
			\For{$j=1,\dots,i$}{
				Form quadratic function $Q^{ij}(x) = x^i x^j$

				Compute $\Sigma^{ij} = \left(\Aop^T Q^{ij}\right)^* / \alpha - \mu^i\cdot \mu^j$

				Set $\Sigma^{ji} = \Sigma^{ij}$

@@

	}
	\caption{Compute scaling factor $\alpha$, mean $\mu$, and covariance $\Sigma$}
	\label{alg:varhpi_mean_cov}
\end{algorithm2e}

@@

	\Input{Finite set of candidate points $P\subset \Omega$, \\mean $\mu(x)$ and covariance $\Sigma(x)$, \\domain boundary information $\partial \Omega$, \\previous sample point batches $S_1, \dots, S_{b-1}$}
	\Output{Batch of new sample points $S_b$}

@@

			Remove all points $p'$ satisfying $\mu(p') \in E_p$ from $P$
		}
	}

@@

First, we use fast H-matrix addition to symmetrize $\AmatPc$:
\begin{equation*}
	\AmatPcSym := \frac{1}{2}\left(\AmatPc + \AmatPc^T\right).
\end{equation*}
Then we modify $\AmatPcSym$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
	\label{eq:rational_matrix_function}
	\AmatPcSymPlus := c_0 \mathbf{I} + c_1 \AmatPcSym + c_2 \left(\AmatPcSym + \mu \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\AmatPcSym$. Approximation \eqref{eq:rational_matrix_function} may be written as $\AmatPcSymPlus = \ratfct(\AmatPcSym)$, where $\ratfct$ is the following rational function:
\begin{equation}
	\label{eq:rational_function_scalar}
	\ratfct(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\mu}.
\end{equation}
The scalars $c_0$, $c_1$, $c_2$, and $\mu$ are chosen so that the moderate and large positive eigenvalues of $\AmatPcSymPlus$ approximately equal the corresponding moderate and large positive eigenvalues of $\AmatPcSym$, while the negative and small positive eigenvalues of $\AmatPcSym$ are modified so that $\AmatPcSymPlus$ is positive semi-definite. Later in this section we will explain how $c_0$, $c_1$, $c_2$, and $\mu$ are chosen and how our choice of these scalars affects the spectral properties of $\AmatPcSymPlus$. In Figure FIG we illustrate $\ratfct(\lambda)$. Forming $\AmatPcSymPlus$ requires computing an H-matrix inverse, $\left(\AmatPcSym + \mu \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is computationally acceptable because $\AmatPcSym + \mu \mathbf{I}$ is an H-matrix. Inversion of H-matrices is a routine procedure which only requires $O(N \log(N)^2)$ work \cite{Hmatrixinverse}.

We choose $c_0, c_1, c_2$, and $\mu$, to be the solution to the following system of equations:
\begin{subequations}
	\label{eq:four_eq_system}
	\begin{align}
		\ratfct(0) &= 0 \label{eq:desired_property_1} \\
		\ratfct'(0) &= 0 \label{eq:desired_property_2} \\
		\ratfct(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
		\ratfct(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
	\end{align}
\end{subequations}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\AmatPcSym$, respectively. We also require $\mu > |\lambda_\text{min}|$, so that the pole of $\ratfct$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. Note that $\lambda_\text{min} < 0$ since $\AmatPcSym$ is indefinite. 
We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}. We describe how we solve system \eqref{eq:four_eq_system} in Appendix \ref{app:solve_rational_system}. Solving \eqref{eq:four_eq_system} reduces to solving a smooth one-dimensional optimization problem, which is computationally trivial.

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\mu > |\lambda_\text{min}|$, imply that $\ratfct(0)=0$ is the unique minimum of $\ratfct$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $\ratfct$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $\AmatPcSymPlus$ is positive semi-definite. Furthermore, \eqref{eq:desired_property_1} implies that the null space of $\AmatPcSymPlus$ contains the null-space of $\AmatPcSym$. This is important when the operator $\Aop$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. Equation \eqref{eq:desired_property_4} forces the moderate and large positive eigenvalues of $\AmatPcSymPlus$ to be close to the corresponding moderate and large positive eigenvalues of $\AmatPcSym$. I.e., $\ratfct$ modifies the ``important'' part of the spectrum of $\AmatPcSym$ as little as possible. Equation \eqref{eq:desired_property_3} ensures that the negative eigenvalues of $\AmatPcSym$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $\ratfct$. There is a tradeoff: if we choose a larger value for $s(\lambda_\text{min})$, then the positive eigenvalues of $\AmatPcSymPlus$ will be closer to the corresponding positive eigenvalues of $\AmatPcSym$ (good), but the erroneous negative eigenvalues of $\AmatPcSym$ will be amplified in magnitude (bad). This is shown in Figure FIG. We find that $\ratfct(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.

@@

\section{Fast ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_p$ and $E_q$ be the ellipsoids defined as
\begin{align*}
	E_p :=& \{x : (x - \mu_p)^T \Sigma_p^{-1} (x - \mu_p) \le \tau^2\} \\
	E_q :=& \{x : (x - \mu_q)^T \Sigma_q^{-1} (x - \mu_q) \le \tau^2\}, \\
\end{align*}
where $\mu_p, \mu_q \in \mathbb{R}^d$, and $\Sigma_p, \Sigma_q \in \mathbb{R}^{d \times d}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\mu_p - \mu_q)^T \left(\frac{1}{1-s}\Sigma_p + \frac{1}{s}\Sigma_q\right)^{-1}(\mu_p - \mu_q)	
\end{equation*}
In \cite{ELLIPSOIDINTERSECT} it is shown that $E_p \cap E_q = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$. We check whether $E_p$ and $E_q$ intersect by minimizing $K(s)$ on $(0,1)$. If $K(s^*) <0$ at the minimizer $s^*$, then $E_p \cap E_q = \{\}$. Otherwise $E_p \cap E_q \neq \{\}$.

The function $K(s)$ may be evaluated quickly for many $s$ by pre-computing the solution to the generalized eigenvalue problem
\begin{equation*}
	\Sigma_p P = \Sigma_q P \Lambda,
\end{equation*}
where $P \in \mathbb{R}^{d \times d}$ is the matrix of generalized eigenvectors (which may be non-orthogonal), and $\Lambda = \diag(\lambda_1,\lambda_2,\dots,\lambda_d)$ is the diagonal matrix of generalized eigenvalues $\lambda_i$. The matrix $P$ simultaneously diagonalizes $\Sigma_p$ and $\Sigma_q$, in the sense that $P^T\Sigma_p P = \Lambda$, and $P^T\Sigma_q P = I$, where $I$ is the $d \times d$ identity matrix. Using this diagonalization, and some algebraic manipulations, we may write $K(s)$ as
\begin{equation}
	\label{eq:Ks_generalized}
	K(s) = 1 - \frac{1}{\tau^2} \sum_{i=1}^d \frac{s(1-s)}{1 + s (\lambda_i - 1)}v_i^2,
\end{equation}
where $v := \Phi^T\left(\mu_p - \mu_q\right)$. We compute the generalized eigenvalue decomposition of $\Sigma_p$ and $\Sigma_q$, then minimize $K(s)$ in the form \eqref{eq:Ks_generalized} on the interval $(0,1)$ using Brent's algorithm \cite{BRENT} (any fast 1 dimensional convex optimization routine may be used). The resulting algorithm for checking whether $E_p$ and $E_q$ intersect is summarized in Algorithm \ref{alg:ellipsoid_intersection_test}.

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	\SetKwProg{Fn}{Function}{}{}
	\Input{Ellipsoid $E_p$ with mean $\mu_p$ and covariance $\Sigma_p/\tau^2$\\
		Ellipsoid $E_q$ with mean $\mu_q$ and covariance $\Sigma_q/\tau^2$ 
	}

@@

	Solve generalized eigenvalue problem $\Sigma_p \Phi = \Sigma_q \Phi \Lambda$

	$v \gets \Phi^T\left(\mu_p - \mu_q\right)$

@@

\begin{defn}[Moments of $\impulseresponse_x$]
	\label{defn:alpha_rho_mu_sigma}
	We define the spatially varying scaling factor, $\alpha$, the scaled impulse response, $\widehat{\impulseresponse}_x$, the spatially varying mean, $\mu$, and the spatially varying covariance, $\Sigma$, as follows:
	\begin{align*}
		\alpha:\Omega \rightarrow \mathbb{R}, &\qquad \alpha(x) := \int_{\Omega} \impulseresponse_x(z) dz, \\
		\widehat{\impulseresponse}_x:\Omega \rightarrow \mathbb{R}, &\qquad \widehat{\impulseresponse}_x := \impulseresponse_x \big/ \alpha(x), \\
		\mu:\Omega \rightarrow \mathbb{R}^d, &\qquad \mu(x) := \int_\Omega z \widehat{\impulseresponse}_x(z) dz, \\
		\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}, &\qquad\Sigma(x) := \int_\Omega (z - \mu(x))(z - \mu(x))^T \widehat{\impulseresponse}_x(z) dz.
	\end{align*}
\end{defn}

\begin{proof}[Proof of Theorem \ref{thm:vol_mean_cov}]
	Let $v \in L^2$ be arbitrary. We have
	\begin{align*}
		\left(v, \alpha\right)_{L^2(\Omega)} =& \int_\Omega v(x) \int_{\Omega} \left(\Aop \delta_x^*\right)^*(z) dz dx\\
		=& \int_\Omega v(x) \int_{\Omega} A(z,x) dz dx \\
		=& \int_\Omega \int_\Omega v(x) A(z,x) C(z) dz dx \\
		=& \left(\Aop^T C\right)(v),
	\end{align*}
	which implies $\alpha = \left(\Aop^T C\right)^*$.

>>>>

		Compute $\spatialvol$, $\spatialmean$, and $\spatialcov$ using Algorithm \ref{alg:varhpi_mean_cov}.

@@

Because of the non-negative kernel property, for each $x \in \Omega$, the impulse response $\impulseresponse_x$ is a non-negative function $y \mapsto \impulseresponse_x(y)$. Hence, $\impulseresponse_x$ is a scaled probability distribution. Let $\spatialvol$ denote the scaling factor, let $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denote the normalized version of $\impulseresponse_x$, and let $\spatialmean(x)$ and $\spatialcov(x)$ denote the mean and covariance of $\widehat{\impulseresponse}_x$, respectively. We make the approximation that the support of $\impulseresponse_x$ is contained within the ellipsoid
\begin{equation}
	\label{eq:support_ellipsoid}
	E_x := \{x' \in \Omega: (x' - \spatialmean(x))^T \spatialcov(x)^{-1} (x' - \spatialmean(x)) \le \tau^2\},
\end{equation}
where $\tau$ is a fixed constant. The ellipsoid $E_x$ is the set of points within $\tau$ standard deviations from the mean of the Gaussian distribution with mean $\spatialmean(x)$ and covariance $\spatialcov(x)$, i.e., the Gaussian distribution which has the same mean and covariance as $\widehat{\impulseresponse}_x$ (see Figure FIG). The quantity $\tau$ is a parameter that must be chosen by the user of the algorithm. The larger $\tau$ is, the larger the ellipsoid $E_x$ is, and the more conservative the estimate is for the support of $\impulseresponse_x$. However, the cost of our algorithm will depend on how many non-overlapping ellipsoids $E_x$ we can ``pack'' in the domain $\Omega$ (more ellipsoids $\implies$ lower cost), and choosing a larger value of $\tau$ means that fewer ellipsoids will fit in $\Omega$. We find that $\tau=3$ yields a reasonable balance between these competing interests, and use $\tau=3$ in our numerical results. The fraction of the ``mass'' of $\impulseresponse_x$ residing outside of $E_x$ is less than $1/\tau^2$ by Chebyshev's inequality, though this bound is conservative and typically far less mass resides in this region.

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\spatialvol(x)$, the length-$d$ vector $\spatialmean(x)$, and the $d \times d$ matrix $\spatialcov(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
	Let $\spatialvol:\Omega \rightarrow \mathbb{R}$, $\spatialmean:\Omega \rightarrow \mathbb{R}^d$, and $\spatialcov:\Omega \rightarrow \mathbb{R}^{d \times d}$, be the following functions:
	\begin{equation*}
		\spatialvol(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
		\spatialmean(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
		\spatialcov(x) := \Var(\widehat{\impulseresponse}_x),
	\end{equation*}
	where $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^d\right)$. Let $C$, $\{L^i\}_{i=1}^d$, and ${\{Q^{ij}\}_{i=1}^d}_{j=1}^d$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
		C(x) := 1, \qquad
		L^i(x) := x^i, \qquad
		Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ to denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ to denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
			\spatialvol =& \left(\Aop^T C\right)^* \\
			\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
			\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
		\end{align}
	\end{subequations}
	 for $i=1,\dots, d$, $j=1,\dots,d$.
\end{theorem}

@@

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	{	
		\Input{Operator $\Aop$}
		\Output{$\spatialvol$, $\spatialmean$, and $\spatialcov$}

		\tcp{Compute scaling factor $\spatialvol$}

@@

		Compute $\spatialvol = \left(\Aop^T C\right)^*$

		\tcp{Compute mean $\spatialmean$}
		\For{$i=1,2,\dots,d$}{
			Form linear function $L^i(x) = x^i$

			Compute $\spatialmean^i = \left(\Aop^T L^i\right)^* / \spatialvol$
		}
		\tcp{Compute covariance $\spatialcov$}
		\For{$i=1,2,\dots,d$}{
			\For{$j=1,\dots,i$}{
				Form quadratic function $Q^{ij}(x) = x^i x^j$

				Compute $\spatialcov^{ij} = \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j$

				Set $\spatialcov^{ji} = \spatialcov^{ij}$

@@

	}
	\caption{Compute scaling factor $\spatialvol$, mean $\spatialmean$, and covariance $\spatialcov$}
	\label{alg:varhpi_mean_cov}
\end{algorithm2e}

@@

	\Input{Finite set of candidate points $P\subset \Omega$, \\spatialmean $\spatialmean(x)$ and covariance $\spatialcov(x)$, \\domain boundary information $\partial \Omega$, \\previous sample point batches $S_1, \dots, S_{b-1}$}
	\Output{Batch of new sample points $S_b$}

@@

			Remove all points $p'$ satisfying $\spatialmean(p') \in E_p$ from $P$
		}
	}

@@

First, we use fast H-matrix addition to symmetrize $\AmatPc$:
\begin{equation*}
	\AmatPcSym := \frac{1}{2}\left(\AmatPc + \AmatPc^T\right).
\end{equation*}
Then we modify $\AmatPcSym$ to make it positive semi-definite by forming a rational matrix function of the following form: 
\begin{equation}
	\label{eq:rational_matrix_function}
	\AmatPcSymPlus := c_0 \mathbf{I} + c_1 \AmatPcSym + c_2 \left(\AmatPcSym + \ratpole \mathbf{I}\right)^{-1},
\end{equation}
where $\mathbf{I}$ is the identity matrix which has the same shape as $\AmatPcSym$. Approximation \eqref{eq:rational_matrix_function} may be written as $\AmatPcSymPlus = \ratfct(\AmatPcSym)$, where $\ratfct$ is the following rational function:
\begin{equation}
	\label{eq:rational_function_scalar}
	\ratfct(\lambda) := c_0 + c_1 \lambda + \frac{c_2}{\lambda+\ratpole}.
\end{equation}
The scalars $c_0$, $c_1$, $c_2$, and $\ratpole$ are chosen so that the moderate and large positive eigenvalues of $\AmatPcSymPlus$ approximately equal the corresponding moderate and large positive eigenvalues of $\AmatPcSym$, while the negative and small positive eigenvalues of $\AmatPcSym$ are modified so that $\AmatPcSymPlus$ is positive semi-definite. Later in this section we will explain how $c_0$, $c_1$, $c_2$, and $\ratpole$ are chosen and how our choice of these scalars affects the spectral properties of $\AmatPcSymPlus$. In Figure FIG we illustrate $\ratfct(\lambda)$. Forming $\AmatPcSymPlus$ requires computing an H-matrix inverse, $\left(\AmatPcSym + \ratpole \mathbf{I}\right)^{-1}$. While explicit computation of matrix inverses should usually be avoided, here it is computationally acceptable because $\AmatPcSym + \ratpole \mathbf{I}$ is an H-matrix. Inversion of H-matrices is a routine procedure which only requires $O(N \log(N)^2)$ work \cite{Hmatrixinverse}.

We choose $c_0, c_1, c_2$, and $\ratpole$, to be the solution to the following system of equations:
\begin{subequations}
	\label{eq:four_eq_system}
	\begin{align}
		\ratfct(0) &= 0 \label{eq:desired_property_1} \\
		\ratfct'(0) &= 0 \label{eq:desired_property_2} \\
		\ratfct(\lambda_\text{min}) &= |\lambda_\text{min}| \label{eq:desired_property_3} \\
		\ratfct(\lambda_\text{max}) &= \lambda_\text{max}, \label{eq:desired_property_4}
	\end{align}
\end{subequations}
where $\lambda_\text{min}$ and $\lambda_\text{max}$ are the smallest and largest eigenvalues of $\AmatPcSym$, respectively. We estimate $\lambda_\text{min}$ and $\lambda_\text{max}$ using the implicitly restarted Lanczos method \cite{lanczos,scipy}. We assume that $\lambda_\text{min} < 0$; if $\lambda_\text{min} \ge 0$, then $\AmatPcSym$ is already positive semi-definite, so we do not need to modify it with a rational function. We also require $\ratpole > |\lambda_\text{min}|$, so that the pole of $\ratfct$ is not in $[\lambda_\text{min},\lambda_\text{max}]$. We describe how we solve system \eqref{eq:four_eq_system} in Appendix \ref{app:solve_rational_system}. Solving \eqref{eq:four_eq_system} reduces to solving a smooth one-dimensional optimization problem, which is computationally trivial.

Together, equations \eqref{eq:desired_property_1}, \eqref{eq:desired_property_2}, \eqref{eq:desired_property_3},  and \eqref{eq:desired_property_4}, and the condition $\ratpole > |\lambda_\text{min}|$, imply that $\ratfct(0)=0$ is the unique minimum of $\ratfct$ on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that $\ratfct$ is non-negative on $[\lambda_\text{min},\lambda_\text{max}]$, which implies that  $\AmatPcSymPlus$ is positive semi-definite. Furthermore, \eqref{eq:desired_property_1} implies that the null space of $\AmatPcSymPlus$ contains the null-space of $\AmatPcSym$. This is important when the operator $\Aop$ has a large (or infinite) dimensional null space, or a spectrum that clusters at zero. Hessians in ill-posed distributed parameter inverse problems typically have both of these properties. Equation \eqref{eq:desired_property_4} forces the moderate and large positive eigenvalues of $\AmatPcSymPlus$ to be close to the corresponding moderate and large positive eigenvalues of $\AmatPcSym$. I.e., $\ratfct$ modifies the ``important'' part of the spectrum of $\AmatPcSym$ as little as possible. Equation \eqref{eq:desired_property_3} ensures that the negative eigenvalues of $\AmatPcSym$, which result from error in the product-convolution approximation, do not get amplified in magnitude by the function $\ratfct$. There is a tradeoff: if we choose a larger value for $s(\lambda_\text{min})$, then the positive eigenvalues of $\AmatPcSymPlus$ will be closer to the corresponding positive eigenvalues of $\AmatPcSym$ (good), but the erroneous negative eigenvalues of $\AmatPcSym$ will be amplified in magnitude (bad). This is shown in Figure FIG. We find that $\ratfct(\lambda_\text{min})=|\lambda_\text{min}|$ is an appropriate balance between these competing interests.

@@

\section{Fast ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_p$ and $E_q$ be the ellipsoids defined as
\begin{align*}
	E_p :=& \{x : (x - \spatialmean_p)^T \spatialcov_p^{-1} (x - \spatialmean_p) \le \tau^2\} \\
	E_q :=& \{x : (x - \spatialmean_q)^T \spatialcov_q^{-1} (x - \spatialmean_q) \le \tau^2\}, \\
\end{align*}
where $\spatialmean_p, \spatialmean_q \in \mathbb{R}^d$, and $\spatialcov_p, \spatialcov_q \in \mathbb{R}^{d \times d}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_p - \spatialmean_q)^T \left(\frac{1}{1-s}\spatialcov_p + \frac{1}{s}\spatialcov_q\right)^{-1}(\spatialmean_p - \spatialmean_q)	
\end{equation*}
In \cite{ELLIPSOIDINTERSECT} it is shown that $E_p \cap E_q = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$. We check whether $E_p$ and $E_q$ intersect by minimizing $K(s)$ on $(0,1)$. If $K(s^*) <0$ at the minimizer $s^*$, then $E_p \cap E_q = \{\}$. Otherwise $E_p \cap E_q \neq \{\}$.

The function $K(s)$ may be evaluated quickly for many $s$ by pre-computing the solution to the generalized eigenvalue problem
\begin{equation*}
	\spatialcov_p P = \spatialcov_q P \Lambda,
\end{equation*}
where $P \in \mathbb{R}^{d \times d}$ is the matrix of generalized eigenvectors (which may be non-orthogonal), and $\Lambda = \diag(\lambda_1,\lambda_2,\dots,\lambda_d)$ is the diagonal matrix of generalized eigenvalues $\lambda_i$. The matrix $P$ simultaneously diagonalizes $\spatialcov_p$ and $\spatialcov_q$, in the sense that $P^T\spatialcov_p P = \Lambda$, and $P^T\spatialcov_q P = I$, where $I$ is the $d \times d$ identity matrix. Using this diagonalization, and some algebraic manipulations, we may write $K(s)$ as
\begin{equation}
	\label{eq:Ks_generalized}
	K(s) = 1 - \frac{1}{\tau^2} \sum_{i=1}^d \frac{s(1-s)}{1 + s (\lambda_i - 1)}v_i^2,
\end{equation}
where $v := \Phi^T\left(\spatialmean_p - \spatialmean_q\right)$. We compute the generalized eigenvalue decomposition of $\spatialcov_p$ and $\spatialcov_q$, then minimize $K(s)$ in the form \eqref{eq:Ks_generalized} on the interval $(0,1)$ using Brent's algorithm \cite{BRENT} (any fast 1 dimensional convex optimization routine may be used). The resulting algorithm for checking whether $E_p$ and $E_q$ intersect is summarized in Algorithm \ref{alg:ellipsoid_intersection_test}.

\begin{algorithm2e}
	\SetAlgoNoLine
	\SetKwInOut{Input}{Input}
	\SetKwInOut{Output}{Output}
	\SetKwProg{Fn}{Function}{}{}
	\Input{Ellipsoid $E_p$ with mean $\spatialmean_p$ and covariance $\spatialcov_p/\tau^2$\\
		Ellipsoid $E_q$ with mean $\spatialmean_q$ and covariance $\spatialcov_q/\tau^2$ 
	}

@@

	Solve generalized eigenvalue problem $\spatialcov_p \Phi = \spatialcov_q \Phi \Lambda$

	$v \gets \Phi^T\left(\spatialmean_p - \spatialmean_q\right)$

@@

\begin{defn}[Moments of $\impulseresponse_x$]
	\label{defn:alpha_rho_mu_sigma}
	We define the spatially varying scaling factor, $\spatialvol$, the scaled impulse response, $\widehat{\impulseresponse}_x$, the spatially varying mean, $\spatialmean$, and the spatially varying covariance, $\spatialcov$, as follows:
	\begin{align*}
		\spatialvol:\Omega \rightarrow \mathbb{R}, &\qquad \spatialvol(x) := \int_{\Omega} \impulseresponse_x(z) dz, \\
		\widehat{\impulseresponse}_x:\Omega \rightarrow \mathbb{R}, &\qquad \widehat{\impulseresponse}_x := \impulseresponse_x \big/ \spatialvol(x), \\
		\spatialmean:\Omega \rightarrow \mathbb{R}^d, &\qquad \spatialmean(x) := \int_\Omega z \widehat{\impulseresponse}_x(z) dz, \\
		\spatialcov:\Omega \rightarrow \mathbb{R}^{d \times d}, &\qquad\spatialcov(x) := \int_\Omega (z - \spatialmean(x))(z - \spatialmean(x))^T \widehat{\impulseresponse}_x(z) dz.
	\end{align*}
\end{defn}

\begin{proof}[Proof of Theorem \ref{thm:vol_mean_cov}]
	Let $v \in L^2$ be arbitrary. We have
	\begin{align*}
		\left(v, \spatialvol\right)_{L^2(\Omega)} =& \int_\Omega v(x) \int_{\Omega} \left(\Aop \delta_x^*\right)^*(z) dz dx\\
		=& \int_\Omega v(x) \int_{\Omega} A(z,x) dz dx \\
		=& \int_\Omega \int_\Omega v(x) A(z,x) C(z) dz dx \\
		=& \left(\Aop^T C\right)(v),
	\end{align*}
	which implies $\spatialvol = \left(\Aop^T C\right)^*$.
